---
title: "STAT 692 Final Project: NYC Housing market price modeling by location"
author: "Riana Guha"
date: "Spring 2024"
output:
  html_document: 
    toc: yes
  pdf_document: default
urlcolor: cyan
---

# Introduction

The main aim of this project is to model the NYC housing data to determine the optimal house price. I will also be focusing on the house location and it's importance in the price. The dataset is avaliable on kaggle - https://www.kaggle.com/datasets/nelgiriyewithana/new-york-housing-market/data

Project objectives:
Price Analysis: find the ideal price range for houses using predictive modeling. This will provide insights for investors and real estate agents to make informed decisions before purchasing a house in NYC.

Location Analysis: Analyze the importance of house location (neighborhoods, districts, regions).

Key Features:
-`BROKERTITLE`: Title of the broker
-`TYPE`: Type of the house
-`PRICE`: Price of the house
-`BEDS`: Number of bedrooms
-`BATH`: Number of bathrooms
-`PROPERTYSQFT`: Square footage of the property
-`ADDRESS`: Full address of the house
-`STATE`: State of the house
-`MAIN_ADDRESS`: Main address information
-`ADMINISTRATIVE_AREA_LEVEL_2`: Administrative area level 2 information
-`LOCALITY`: Locality information
-`SUBLOCALITY`: Sublocality information
-`STREET_NAME`: Street name
-`LONG_NAME`: Long name
-`FORMATTED_ADDRESS`: Formatted address
-`LATITUDE`: Latitude coordinate of the house
-`LONGITUDE`: Longitude coordinate of the house


Loading the required packages
```{r}
require(tidyverse)
```


Data exploration and cleaning
```{r}
data <- read.csv("NY-House-Dataset.csv")
str(data) # structure of dataset
dim(data) # the dataset contains 4801 rows and 17 columns
colSums(is.na(data)) # checking for missing values per column - No missing values

head(data) # viewing the first 6 rows

data_filtered = data[,c('TYPE', 'PRICE', 'BEDS', 'BATH', 'STATE', 'PROPERTYSQFT', 'ADMINISTRATIVE_AREA_LEVEL_2', 'LOCALITY', 'SUBLOCALITY', 'STREET_NAME', 'LONG_NAME', 'LATITUDE', 'LONGITUDE')] # filtering data to keep relevant columns
colnames(data_filtered) # verifying column names
dim(data_filtered) # it is now a 4801x13 dataset
head(data_filtered)
```


EDA (Exploratory Data Analysis)
```{r}

```

